MLH1 data report 

### 2017-09-25 rmd_report.rmd
updated report md file

TODO
remove the output for the plots
data set up page; remove non quality observations
Solve the cells per mouse problem
figure out figure captions
add a place to record the 'batches' that make up the data
category has CAST female, this might be messing things up, since there is not CAST data

include is for blocking output messages. echo false is for silencing the code commands.
```{r setup, echo=FALSE, include = FALSE}
library(knitr)
library(ggplot2)
library(pwr)
library(plyr)
library(lattice)
library(dplyr)
#setwd("C:/Users/alpeterson7/Documents/MLH1repo")
load(file="MLH1_data_setup.RData")

@@

#### goals of research
Measure nMLH1 foci per meiotic cell to estimate recombination rate for diverse strains of house mice (rodents). Comparisons the differences in recombination rates across sexes and genetic background to inform models of how meiotic recombination rates evolve. (insert small picture of meixyte)

### Goal / Discription of Report
Goal of this report;

> *to keep track of the number of images for each category and mouse*

This is important for getting an idea of the 'power' I have to accurately estimate MLH1 distributions across the categories of cells I am quantifying.

#### Discription of data and quantification process
data frames are loaded of MLH1 data 
imagename, mouse name, strain, sex, nMLH1.

quanti process. anonymized, given quality score

@@

#Add fake CAST female data

last_row <-data.frame(file.name=c("12dec18_20dec20_CAST_f1_sp1_12_1_rev.tif", "12dec18_20dec20_CAST_f1_sp1_12_2_rev.tif"), Random.Name = c("1234567.tif","123456789.tif"), quality=c(1,1),nMLH1.foci =c(25,22), XY.paired =c("no","no"), REDO.crop =c("no","no"),n =c(20,20), achiasmate=c(0,0),asynased=c(0,0),notes=c("",""),category = c("CAST female","CAST female"), strain= c("CAST","CAST"), sex=c("female","female"),adj_nMLH1.foci=c(25,22),mouse=c("20dec20_CAST_f1","20dec20_CAST_f1") )

@@

@@

Use these figures to identify outliers and if there is variance in the means and variance.

```{r first boxplots, echo=FALSE, fig.caption="boxplots of MLH1 distributions", fig.height=5, fig.width=5, fig.align='center'}

ff <- ggplot(MLH1_data, aes(x = mouse, y = nMLH1.foci)) + geom_boxplot(data = MLH1_data, aes(fill = factor(strain) ) )+ theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + scale_fill_manual(values=c("#56B4E9", "cadetblue", "lightblue",
  "coral1", "#E69F00", "yellowgreen"))
ff <- ff + facet_wrap(~ sex)

### Tests for quality and nMLH1 foci number

@@

Boxplots of mice distributions with quality score of 1

```{r high qual, fig.caption="boxplots of MLH1 distributions", fig.height=5, fig.width=5, echo=FALSE}

qq <- ggplot(highest_qual_df, aes(x = mouse, y = nMLH1.foci)) + geom_boxplot(data = highest_qual_df, aes(fill = factor(strain) ) ) +  ggtitle("Cells with quality scores of 2 or 1") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + scale_fill_manual(values=c("#56B4E9", "cadetblue", "lightblue",
  "coral1", "#E69F00", "yellowgreen"))
high_qual_boxplot <- qq + facet_wrap(~ sex) #this is breaking the kniting...

The spacing in these boxplots is off because of 'ghost' cells, which are in the data frame, but not in the subsetted data. (Iguess)

Table for assessing the number of passed mice
```{r, passing mice, echo=FALSE}
q_cutoff_table <- ddply(MLH1_data, .(mouse), summarise,
                        total =  length(nMLH1.foci),
                        q5 = sum(as.numeric(quality) >= 4, na.rm = TRUE ), 
                        above5 = sum(as.numeric(quality) <= 4, na.rm = TRUE )
                       # q_l3 = sum(as.numeric(as.numeric(quality)) <= 4, na.rm = TRUE )
)
#q_cutoff_table

mice_25cells <- q_cutoff_table[q_cutoff_table$above5 > 25,]$mouse
#12 mice have more than 25 cells
mice_20cells <- q_cutoff_table[q_cutoff_table$above5 > 20,]$mouse
#26 mice have over 20 cells
mice_15cells <- q_cutoff_table[q_cutoff_table$above5 > 14,]$mouse

passed_mice <- as.vector(q_cutoff_table[q_cutoff_table$above5 > 15,]$mouse)

#mnake dataframe with just passing mice -- so that it can be ploted
passed_mice_df <- MLH1_data[ ( MLH1_data$quality < 5), ]

#### Boxplots of 'passing' mice distributions

```{r, echo=FALSE}
pp <- ggplot(passed_mice_df, aes(x = mouse, y = nMLH1.foci)) + geom_boxplot(data = passed_mice_df, aes(fill = factor(strain) ) )+ theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + scale_fill_manual(values=c("#56B4E9", "cadetblue", "lightblue",
  "coral1", "#E69F00", "yellowgreen"))
passed_mice_plot <- pp + facet_wrap(~ sex)

```
```{r,  show boxplots, fig.height=5, fig.width=8, echo=FALSE}

ff
high_qual_boxplot
passed_mice_plot
```

@@

Human quantification seems to be biased towards rating cells with more MLH1 foci as higher quality. Unbiased cell quality assignment, would not show a positive correlation with quality and nMLH1.

@@

Potential solutions?

> *do some mice have better quality cells?* *Does this cause bias in the data?*

```{r, fig.height=4, fig.width=4, echo=FALSE, fig.caption= "histogrms", warning=FALSE}
# do the same thing as above but geom_hist

It seems like there is a population of lower cells in MSM males. This could be from slides which had bad staining.

#### Assessing the number of cells per mouse

latice plot of scatter plots for jitter plots of cell oberservations by quality
```{r, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
#TODO, add a line for mouse and strain

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE }
#geom_hline(aes(yintercept = med, group = gr), colour = 'red')
for (gg in 1:length(plot_list) ){
  show(plot_list[gg])
}
```

Making all of these scatter plots, allows us to look at the whole distributions. In some mice there seems to be bimodel cell populations, those of clearly high quality and those of lower quality. This may be due to the screeign of images before quantification, and allowing more bad cells data remain in the data set, instead of deleting them.

@@

Think about a loop that will -- i) take ~100 different sized samples of total images (10%, 25%, 50%, 75%, 90%) (with varying degrees of quality)
ii) then run t.tests across those samples, and report the p value
iii) make a table of the p values for these sample permutations

Diminishing return analysis. Take some of the mice identified above, simulate random sampling of 10,15,20,30. Display the raw mean and sd, and results from 10 simulations.
```{r, comparing sample sizes, echo=FALSE}
PWD_m_example = MLH1_data[MLH1_data$mouse == '10mar15_PWD_m2', ]

smp10dist = sample(PWD_m_example$nMLH1.foci, 10, replace=FALSE)
samplesize = rep("10", 10)
smp10 <- data.frame(smp10dist, samplesize)

smp15dist = sample(PWD_m_example$nMLH1.foci, 15, replace=FALSE)
samplesize = rep("15", 15)
smp15 <- data.frame(smp15dist, samplesize)

smp20dist = sample(PWD_m_example$nMLH1.foci, 20, replace=FALSE)
samplesize = rep("20", 20)
smp20 <- data.frame(smp20dist, samplesize)

smp30dist = sample(PWD_m_example$nMLH1.foci, 30, replace=FALSE)
samplesize = rep("30", 30)
smp10 <- data.frame(smp30dist, samplesize)

#this mouse has 40 cells 
#make a dataframe (so that I can finally make boxplots)
dat <- data.frame(
  MLH1_sample = c(sample(PWD_m_example$adj_nMLH1.foci, 10, replace=FALSE),  sample(PWD_m_example$adj_nMLH1.foci, 15, replace=FALSE),    sample(PWD_m_example$adj_nMLH1.foci, 20, replace=FALSE), sample(PWD_m_example$adj_nMLH1.foci, 30, replace=FALSE), sample(PWD_m_example$adj_nMLH1.foci, 35, replace=FALSE), sample(PWD_m_example$adj_nMLH1.foci, 40, replace=FALSE)),   
  samplesize = c(rep("10", 10), rep("15", 15), rep("20", 20), rep("30", 30), rep("35", 35), rep("40", 40)) )

bb <- ggplot(data = dat, aes(x=samplesize, y=MLH1_sample)) + #geom_boxplot(aes(fill=samplesize))

```{r}
G_f_example = MLH1_data[MLH1_data$mouse == '17mar16_G_f1', ]

datf <- data.frame(
  MLH1_sample = c(sample(G_f_example$nMLH1.foci, 10, replace=FALSE),  sample(G_f_example$nMLH1.foci, 15, replace=FALSE),    sample(G_f_example$nMLH1.foci, 20, replace=FALSE), sample(G_f_example$nMLH1.foci, 30, replace=FALSE), sample(G_f_example$nMLH1.foci, 35, replace=FALSE) ),   
  samplesize = c(rep("10", 10), rep("15", 15), rep("20", 20), rep("30", 30), rep("35", 35)) )

v <- ggplot(data = datf, aes(x=samplesize, y=MLH1_sample)) + geom_boxplot(aes(fill=samplesize) ) 
#v                                                                         
#v + ggtitle("varying sample sizes") 
```
Can the above code be repeated for a larger number of mice (both passing and non-passing) to show the different effects across samples sizes?
power calculations.... from a preivous document --- I showed myself that for 15 (good) cells, would provide a sample big enough to detect a difference of 3 foci. I only did this for a couple mice, I'll try to re-code for every mouse

```{r}
passed_mice_data <- MLH1_data[MLH1_data$mouse %in% passed_mice, ]
non_passed_mice_data <- MLH1_data[MLH1_data$mouse %in% not_passed, ]

@@

The code chunck below makes a table of t.test results. Each category has 2 rows. When I tried changing equal variance to false, the code breaks.
 c("strain", "sex"),

```{r}
#for the record, this is a way to make a table with pwr and t.test results
category_pwr_stat <- ddply(MLH1_data, "category", function(x) {
Nmice <- length( unique(x$mouse) )
mean.count <- mean(x$adj_nMLH1.foci)
sd.count <- sd(x$adj_nMLH1.foci)
cv <- sd.count/mean.count

ttest <- t.test(x$adj_nMLH1.foci, (20), alternative = "two", var.equal=TRUE)$p.value
ttestCI.lowr <- t.test(x$adj_nMLH1.foci, (20), alternative = "two", var.equal=TRUE)$conf.int[1]
ttestCI.upr <- t.test(x$adj_nMLH1.foci, (20), alternative = "two", var.equal=TRUE)$conf.int[2]
ttest.est <- t.test(x$adj_nMLH1.foci, (20), alternative = "two", var.equal=TRUE)$estimate

pwr5mice <- pwr.norm.test(d = (3/sd.count), n = 20, sig.level = 0.05, alternative = "two")$power
data.frame(Nmice=Nmice, cv.count = cv, ttest=ttest, CIlower = ttestCI.lowr, 
           CIupr = ttestCI.upr,
 pwr = pwr5mice, ttest.est=ttest.est)
})
#category_pwr_stat
```

I am not sure how to interpert these power calqs. The estimates are all overlapping.. I guess I shouldn't be suprised by this?

```{r, fig.height=2, fig.width=4, echo=FALSE}
#use kable to make tables. warning this might not work for github Rmd.
kable(AP_strain_table)
```

>>>>

```{r setup, echo=FALSE, include = FALSE}
library(knitr)
library(ggplot2)
library(pwr)
library(plyr)
library(lattice)
library(dplyr)
#setwd("C:/Users/alpeterson7/Documents/MLH1repo")
load(file="MLH1_data_setup.RData")

@@

#### Research Discription
Measure nMLH1 foci per meiotic cell to estimate recombination rate for diverse strains of house mice (rodents). Comparisons the differences in recombination rates across sexes and genetic background to inform models of how meiotic recombination rates evolve. (insert small picture of meixyte)

#### Discription of Report

- Display plots and figures of distributions of MLH1 counts to assess intial overall patterns.

-  *Report basic statistics of results and to help keep track of the number of images for each category and mouse*

#### Discription of quantification process

Aquired microscope cell images are quantified in batches after being anonimzed. The number of MLH1 foci, number of "achiasmate" and asynasped bivalents are quantified. A quality score is given; 1 to 5 (best to worst).

@@

#Add fake CAST female data
last_row <-data.frame(file.name=c("12dec18_20dec20_CAST_f1_sp1_12_1_rev.tif", "12dec18_20dec20_CAST_f1_sp1_12_2_rev.tif"), Random.Name = c("1234567.tif","123456789.tif"), quality=c(1,1),nMLH1.foci =c(25,22), XY.paired =c("no","no"), REDO.crop =c("no","no"),n =c(20,20), achiasmate=c(0,0),asynased=c(0,0),notes=c("",""),category = c("CAST female","CAST female"), strain= c("CAST","CAST"), sex=c("female","female"),adj_nMLH1.foci=c(25,22),mouse=c("20dec20_CAST_f1","20dec20_CAST_f1") )

@@

#### Mice table

Table of the number of mice used and MLH1 stats. (made with kable).

```{r, fig.height=2, fig.width=4, echo=FALSE}
#this table is saved in the .RData file
#use kable to make tables. warning this might not work for github Rmd.
kable(AP_strain_table)
```

@@

```{r first boxplots, echo=FALSE, fig.align='center'}

ff <- ggplot(MLH1_data, aes(x = as.factor(mouse), y = adj_nMLH1.foci)) + geom_boxplot(data = MLH1_data, aes(fill = factor(strain) ) ) + ggtitle("Boxplots of MLH1 distributions by mouse") + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + scale_fill_manual(values=c("#56B4E9", "cadetblue", "lightblue",  "coral1", "#E69F00", "yellowgreen"))
ff <- ff + facet_wrap(~ sex, scales="free")
ff
```

After taking the data from 2 highest cell quality, some but not all mouse means converge.

```{r, histogram, echo=FALSE, fig.caption= "histograms", warning=FALSE, fig.align='center'}

```{r high qual, fig.caption="boxplots of MLH1 distributions", fig.height=5, fig.width=5, echo=FALSE, warnings=FALSE}

qq <- ggplot(highest_qual_df, aes(x = mouse, y = adj_nMLH1.foci)) + geom_boxplot(data = highest_qual_df, aes(fill = factor(strain) ) ) +  ggtitle("Cells with quality scores of 2 or 1") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + scale_fill_manual(values=c("#56B4E9", "cadetblue", "lightblue",
  "coral1", "#E69F00", "yellowgreen"))
high_qual_boxplot <- qq + facet_wrap(~ sex, scales="free") #this is breaking the kniting... 
```

```{r, passing mice, echo=FALSE}
q_cutoff_table <- ddply(MLH1_data, .(mouse), summarise,
                        total =  length(adj_nMLH1.foci),
                        q5 = sum(as.numeric(quality) >= 4, na.rm = TRUE ), 
                        above5 = sum(as.numeric(quality) <= 4, na.rm = TRUE )
                       # q_l3 = sum(as.numeric(as.numeric(quality)) <= 4, na.rm = TRUE )
)
#q_cutoff_table
passed_mice <- as.vector(q_cutoff_table[q_cutoff_table$above5 > 15,]$mouse)

#mnake dataframe with just passing mice -- so that it can be ploted
passed_mice_df <- MLH1_data[ (as.numeric(MLH1_data$quality) < 5), ]
```

#### Comparison of MLH1 distributions

```{r,  show boxplots, fig.height=5, fig.width=8, echo=FALSE}
ff
high_qual_boxplot
```

After taking the data from 2 highest cell quality, some but not all mouse means converge. For some mice (mostly female) the majority of cells are excluded. This is especially true for PWD and WSB females.

#### Tests for quality and nMLH1 foci number

@@

@@

Human quantification seems to be biased towards rating cells with more MLH1 foci as higher quality. Unbiased cell quality assignment, would not show a positive correlation with quality and nMLH1. (CAST female data is fake)

@@

#### Assessing the Distributions by mouse

latice plot of scatter plots for jitter plots of cell oberservations by quality. The category mean is in black and the mouse specific mean is in red.
```{r, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
#TODO, add a line for mouse and strain

@@

```{r, echo=FALSE, message=FALSE,results='hide', warning=FALSE }
#geom_hline(aes(yintercept = med, group = gr), colour = 'red')
for (gg in 1:length(plot_list) ){
  show(plot_list[gg])
  #invisible(lapply(obj,function(x) plot(x,main="some plot")))
}
```

Making all of these scatter plots, allows us to look at the whole distributions of the data for each mouse. The distance of the red line from the black could be a indicator of slides or mice with slide specific technical noise.

PWD females and a very large range. This is driven by data from 17apr15_PWD_f4. These data should be looked at more closely.

In some mice there seems to be bimodel cell populations, those of clearly high quality and those of lower quality. This may be due to the screening of images before quantification, and allowing more bad cells data remain in the data set, instead of deleting them.

@@

Outline discription of the loop for comparing sample size effects

i) take ~100 independant samples of total images across different sample sizes (10%, 25%, 50%, 75%, 90%)
ii) then run t.tests across those samples, and report the p value 
iii) make a table of the p values for these sample permutations
iv) plot the p values or mean p values across sample sizes

Currently this section is under construction.

```{r, include=FALSE}
passed_mice_data <- MLH1_data[MLH1_data$mouse %in% passed_mice, ]
non_passed_mice_data <- MLH1_data[MLH1_data$mouse %in% not_passed, ]

@@
